---
output: pdf_document
---
Examining impact of severe weather events on public health and economy. 
========================================================

## Synopsis

##  Data Processing

In order to load the data into R, the bz2 file is first unzipped using the bunzip2 function in the R.utils package. This package is not part of the standard R package and would need to be installed. In order to prvent an error from bunzip2 on subsequent runs, a check is added to see if the csv file already exists. 

```{r Unzip Data,cache=TRUE, message=FALSE}
if (!file.exists("repdata-data-StormData.csv")){
    install.packages("R.utils")
    library(R.utils)
    bunzip2("repdata-data-StormData.csv.bz2")    
}
```

The data is loaded into a data frame using the read.csv function.
```{r Read Data, cache=TRUE}
storm.data <- read.csv("repdata-data-StormData.csv")
```

Several rows in the storm dataset have a summary of a particular event. These observations don't have any numerical data and are therefore excluded from the analysis. 

``` {r Pre-processing, cache=TRUE}
storm.data <- subset(storm.data,!grepl("summary", storm.data$EVTYPE, ignore.case = TRUE))
```

In order to calculate the total monetry loss, the DMG & DMGEXP fields need to be used. The PROPDMGEXP and CROPEXPDMG fields are expressed in terms of hundreds (h or H), thousands (k or K), millions (m or M) or billions (b or B). All other non numeric values (?,- and +) are replaced by a multiplier of 1. 

The final cash loss is calculated by using the following equation: <br/>

  cash = cropdmg * (10 ^ cropdmpexp) + propdmg * (10 ^ propdmpexp)

```{r Total Cash Loss, cache=TRUE}
pattern <- c("[hH]","[mM]", "[kK]","[bB]", "[-+?]", "^$")
replacement <- c("2","6","3","9","0","0")
for (i in 1:length(pattern)) {
    storm.data$CROPDMGEXP <- sub(pattern[i],replacement[i], storm.data$CROPDMGEXP)
    storm.data$PROPDMGEXP <- sub(pattern[i],replacement[i], storm.data$PROPDMGEXP)    
}

storm.data$CROPDMGEXP <- as.numeric(storm.data$CROPDMGEXP)
storm.data$PROPDMGEXP <- as.numeric(storm.data$PROPDMGEXP)

storm.data$TOTALCASH <- (storm.data$CROPDMG * (10^storm.data$CROPDMGEXP)) + (storm.data$PROPDMG * (10^storm.data$PROPDMGEXP)) 
```

## Exploratory Analysis

In order to study, the impact of different events on population health, the summary function is useful.

```{r summarizing injuries and fatalities}
summary(storm.data$FATALITIES)
summary(storm.data$INJURIES)
```

From the summaries, we see that median & 3rd quartile for both Injuries and Fatalities is zero. However, the max value is large for both the variables. Hence, the data distribution is long tailed. 

To calculate the total impact on human population and economy, we need to sum up the fatalities, injuries & total cash loss for all event types across the dataset. The dataset is melted with to retain only required variables. It is then re-cast to compute the sum for each variable. 

```{r reshaping data,cache=TRUE, message=FALSE}
require(reshape2)
storm.human.melt <- melt(storm.data, id.vars=c("EVTYPE"), measure.vars=c("FATALITIES", "INJURIES", "TOTALCASH"))
storm.dcast.sum <- dcast(storm.human.melt, EVTYPE~variable,sum)
```

The events which have caused the most harm to population health are: 
```{r top events health}
head(storm.dcast.sum[order(storm.dcast.sum$FATALITIES,decreasing=TRUE),c(1,2)],5)
head(storm.dcast.sum[order(storm.dcast.sum$INJURIES,decreasing=TRUE),c(1,3)],5)
```
Clearly, TORNADOS have been the most deadly natural disaster over the past 50 years. 

The events which have caused the most economic damage are:
```{r top events economy}
head(storm.dcast.sum[order(storm.dcast.sum$TOTALCASH,decreasing=TRUE),c(1,4)],5)
```


## Results
<figure>
```{r plots, message=FALSE}
require(ggplot2)
require(gridExtra)
g1 <- ggplot(head(storm.dcast.sum[order(storm.dcast.sum$FATALITIES,decreasing=TRUE),c(1,2)],5), aes(x=EVTYPE, y=FATALITIES)) + geom_bar(stat="identity", fill = "blue") + xlab("Severe Weather Event") + ylab("Total Fatalities")
g2 <- ggplot(head(storm.dcast.sum[order(storm.dcast.sum$INJURIES,decreasing=TRUE),c(1,3)],5), aes(x=EVTYPE, y=INJURIES)) + geom_bar(stat="identity", fill = "blue") + xlab("Severe Weather Event") + ylab("Total Injuries")
#g3 <- ggplot(head(storm.final.sub[order(storm.final.sub$FATALITIES.per.event,decreasing=T),c(1,2,4,5)],5),aes(x=EVTYPE, y=FATALITIES.per.event)) + geom_bar(stat="identity", fill = "green4") 
#g4 <- ggplot(head(storm.final.sub[order(storm.final.sub$INJURIES.per.event,decreasing=T),c(1,3,4,6)],5),aes(x=EVTYPE, y=INJURIES.per.event)) + geom_bar(stat="identity", fill = "green4") 

grid.arrange(g1,g2)
```
<figcaption> Figure 1: Impact of severe weather events on population health </figcaption>
</figure>

It is also interesting to look at the total events recorded per event type.
```{r total events}
EVTYPE.table <- as.data.frame(table(storm.data$EVTYPE))
names(EVTYPE.table) <- c("EVTYPE", "Freq")
storm.final <- merge(storm.dcast.sum, EVTYPE.table)
head(storm.final[order(storm.final$Freq,decreasing = TRUE),],5)
```

From the results, we note that the most frequent events have been HAIL, WIND & TORNADOS. It also interesting to determine the fatalities and injuries per event recorded. To do this, we can divide the total fatalities/injuries by the total events recorded. In the analysis, we have excluded any events that have seen fewer than 20 occurances in the dataset. 

```{r per event, cache=TRUE}
storm.final$FATALITIES.per.event <- storm.final$FATALITIES/storm.final$Freq
storm.final$INJURIES.per.event <- storm.final$INJURIES/storm.final$Freq
storm.final$TOTALCASH.per.event <- storm.final$TOTALCASH/storm.final$Freq
storm.final.sub <- subset(storm.final, storm.final$Freq > 20)
```

The most harmful natural disasters in terms of injuries and fatalities per occurance are:

```{r injuries per event}
head(storm.final.sub[order(storm.final.sub$INJURIES.per.event,decreasing=T),c(1,3,5,7)])
```

```{r fatalities per event}
head(storm.final.sub[order(storm.final.sub$FATALITIES.per.event,decreasing=T),c(1,2,5,6)])
```

The most harmful natural disasters in terms of total cash loss per occurance are:
```{r Total cash per event}
head(storm.final.sub[order(storm.final.sub$TOTALCASH.per.event,decreasing=T),c(1,4,5,8)])
```
